---
title: "R Notebook"
output: html_notebook
---

```{r}
suppressPackageStartupMessages({
library(tidyverse)
library(patchwork)
theme_set(theme_bw())
library(tidymodels)
library(arrow)
})
```

# Import Data:

```{r}
Tor_hitting <- read_csv("TOR_batting.csv")
```

```{r}
NHL_draft <- read_csv("nhldraft.csv")
NHL_draft
```

# Cleaning:

```{r}
NHL_draft_clean <- NHL_draft |>
  select(-id) |>
  mutate(points_per_game = points/games_played,
         position = as.factor(position))

group <- function(data, dataframe, group) {
  data_frame <- data |>
    group_by({{ group }}) |>
    summarise(points = sum(points, na.rm = TRUE),
            games_played = sum(games_played, na.rm = TRUE),
            points_per_game = points/games_played)
}
NHL_draft_clean
```

# Split into training, validation and testing data:

```{r}
NHL_split <- 
  initial_validation_split(NHL_draft_clean, prop = c(0.6, 0.2))
  training <- training(NHL_split)
  validation <- validation(NHL_split)
  test <- testing(NHL_split)
NHL_split
```

```{r}
points_year <- training |>
  group(training, year) |>
  filter(year <= 1996)
points_per_game_until_1996 <- points_year |>
  ggplot(aes(x = year, y = points_per_game)) +
  geom_col() +
  geom_smooth(colour = "red", method = "lm", se = FALSE, formula = y ~ x)
points_per_game_until_1996
points_year
```

# Model Plot 1 and 2

-   **Plot 1 Analysis:**
    -   **Games played as a predictor**: The graph shows how the number of games played correlates with points_per_game, indicating that players with more experience tend to perform better. This suggests that games_played is a relevant feature for predicting player effectiveness.
    -   **Position-based variance**: By colouring the points by position, the plot reveals how the relationship between games_played and points_per_game may vary across different positions. This suggests that the model should account for both games_played and position to improve predictions.
-   **Plot 2 Analysis:**
    -   **Overall pick as a predictor**: The graph shows how a players overall pick correlates with points_per_game, suggesting that lower overall picks (e.g. 1st overall) are likely to perform better. This relationship makes overall_pick a valuable feature for predicting player performance.
    -   **Position-based variance**: By coloring the points by position, the graph highlights potential interactions between overall picks and position type, suggesting the need to include both overall_pick and positon in the model for more accurate predictions.

```{r}
model_plot_1 <- training |>
  ggplot(aes(x = games_played, y = points_per_game, color = position)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(
    title = "Relationship Between Games Played and Points Per Game by Position",
    x = "Games Played",
    y = "Points Per Game",
    color = "Position"
  ) +
  theme_minimal()


model_plot_2 <- training |>
  filter(!is.na(overall_pick)) |>  
  ggplot(aes(x = overall_pick, y = points_per_game, color = position)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed") +
  labs(
    title = "Overall Pick vs. Points Per Game by Position",
    x = "Overall Pick (Lower is Better)",
    y = "Points Per Game",
    color = "Position"
  ) +
  theme_minimal()

model_plot_1
model_plot_2


```
