---
title: "NHL Points Per Game Report"
author: "Group 9"
date: "December 4, 2024"
output: pdf_document
---

```{r}
suppressPackageStartupMessages({
library(tidyverse)
library(patchwork)
theme_set(theme_bw())
library(tidymodels)
library(arrow)
library(stringr)
library(ggplot2)
library(ggridges)
})
```

# Import Data:

```{r}
Tor_hitting <- read_csv("TOR_batting.csv")
```

```{r}
NHL_draft <- read_csv("nhldraft.csv")
NHL_draft
```

# Cleaning:

```{r}
NHL_draft_clean <- NHL_draft |>
  select(-id) |>
  mutate(points_per_game = points/games_played,
         # Change values so that whatever comes first, is what is outputted
         position = str_replace(position, "Centr", "C"), 
         position = str_replace(position, "C; LW", "C"), 
         position = str_replace(position, "C; RW", "C"), 
         position = str_replace(position, "C / R", "C"),
         position = str_replace(position, "C RW", "C"),
         position = str_replace(position, "L/RW", "W"),
         position = str_replace(position, "C/W", "F"),
         position = str_replace(position, "D/C", "D"),
         position = str_replace(position, "D/LW", "D"),
         position = str_replace(position, "D/RW", "D"),
         position = str_replace(position, "LW/C", "LW"),
         position = str_replace(position, "RW/C", "RW"),
         position = str_replace(position, "C/D", "C"),
         position = str_replace(position, "C/RW", "C"),
         position = str_replace(position, "C/LW", "C"),
         position = str_replace(position, "LW/D", "LW"),
         position = str_replace(position, "RW/D", "RW"),
         position = as.factor(position),
         pick_in_round = overall_pick %% 31
  )

group <- function(data, dataframe, group) {
  data_frame <- data |>
    group_by({{ group }}) |>
    summarise(points = sum(points, na.rm = TRUE),
            games_played = sum(games_played, na.rm = TRUE),
            points_per_game = points/games_played)
}
NHL_draft_clean
```

# Split into training, validation and testing data:

```{r}
NHL_split <- 
  initial_validation_split(NHL_draft_clean, prop = c(0.6, 0.2))
  training <- training(NHL_split)
  validation <- validation(NHL_split)
  test <- testing(NHL_split)
NHL_split
```

# Exploratory Plot 1

```{r}
position_by_year <- training |>
  ggplot(aes(x = pick_in_round, y = team)) +
  geom_density_ridges(bandwidth = 1)

position_by_year
```

# Exploratory Plot 2

```{r}
teams_by_draft_pick <- training |>
  ggplot(aes(x = position, y = overall_pick)) +
  geom_boxplot(fill = "aquamarine")

teams_by_draft_pick
```

# Exploratory Plot Analysis

-   **Plot 1:**

    -   Comparing team by pick in round: This graph shows where each team has picked in the draft in the past. This graph helps give us an idea with where teams usually pick in the draft so that if we see that teams have good players, we can connect that to picking lower in the draft rounds.

-   **Plot 2:**

    -   Comparing position to overall pick: This graph shows where each position is picked in the draft generally. It seems that goalies tend to be picked higher and Left Wingers tend to be picked lower in the draft. This can be factored in later as if position tends to be a better position in points per game, we can attribute that back to that left wingers are picked lower generally in the draft.

# Model Plot 1 and 2

-   **Plot 1 Analysis:**
    -   **Games played as a predictor**: The graph shows how the number of games played correlates with points_per_game, indicating that players with more experience tend to perform better. This suggests that games_played is a relevant feature for predicting player effectiveness.
    -   **Position-based variance**: By colouring the points by position, the plot reveals how the relationship between games_played and points_per_game may vary across different positions. This suggests that the model should account for both games_played and position to improve predictions.
-   **Plot 2 Analysis:**
    -   **Overall pick as a predictor**: The graph shows how a players overall pick correlates with points_per_game, suggesting that lower overall picks (e.g. 1st overall) are likely to perform better. This relationship makes overall_pick a valuable feature for predicting player performance.
    -   **Position-based variance**: By coloring the points by position, the graph highlights potential interactions between overall picks and position type, suggesting the need to include both overall_pick and positon in the model for more accurate predictions.

```{r}
model_plot_1 <- training |>
  ggplot(aes(x = games_played, y = points_per_game, color = position)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(
    title = "Relationship Between Games Played and Points Per Game by Position",
    x = "Games Played",
    y = "Points Per Game",
    color = "Position"
  ) +
  theme_minimal()


model_plot_2 <- training |>
  filter(!is.na(overall_pick)) |>  
  ggplot(aes(x = overall_pick, y = points_per_game, color = position)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed") +
  labs(
    title = "Overall Pick vs. Points Per Game by Position",
    x = "Overall Pick (Lower is Better)",
    y = "Points Per Game",
    color = "Position"
  ) +
  theme_minimal()

model_plot_1
model_plot_2


```

# Linear Models:

```{r}
linear_recipe_1 = recipe(points_per_game ~ games_played , data = NHL_draft_clean)
recipe_1_plot=workflow()|>
add_model(linear_reg())|>
add_recipe(linear_recipe_1)|>
fit(training)|>
predict(new_data=validation)|>
bind_cols(validation)|>
ggplot()+
aes(x=games_played,y=.pred)+
geom_point()
recipe_1_plot+
  labs(
    title = "Relationship Between Games Played and Points Per Game",
    x = "Games Played",
    y = "Points Per Game"
  )

linear_recipe_2 = recipe(points_per_game ~ overall_pick  , data = NHL_draft_clean)
recipe_2_plot=workflow()|>
add_model(linear_reg())|>
add_recipe(linear_recipe_2)|>
fit(training)|>
predict(new_data=validation)|>
bind_cols(validation)|>
ggplot()+
aes(x=overall_pick,y=.pred)+
geom_point()
recipe_2_plot+
  labs(
    title = "Overall Pick vs. Points Per Game",
    x = "Overall Pick (Lower is Better)",
    y = "Points Per Game"
  )
NHL_recipes=list(
  basic = linear_recipe_1,
  interact = linear_recipe_2  
)

```

# Final Linear Model:

```{r}

#Setting up workflow
lm_model=linear_reg()|>
set_engine("lm")
NHL_workflow_set=workflow_set(preproc = NHL_recipes, models = list(lm=lm_model))

NHL_workflow_set
#Fitting to the training data
NHL_fit = workflow_map(
  NHL_workflow_set, fn="tune_grid", seed = 15,
  resamples=validation_set(NHL_split)
)

NHL_fit |>
    collect_metrics() |>
    ggplot() +
    aes(x = wflow_id, y = mean) +
    geom_col(fill = "lightgrey", colour = "black") +
    facet_wrap(~ .metric, scales = "free")

```
